Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2019-12-28 article/Old versions/JAS/PCADSC_article_MPS_final.tex

@@

The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}$$
 be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
\begin{enumerate}
\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{\mathbf{X}}_{j} \in \RR^{n_x \times d}$ be the standardized dataset.
\item Form the principal component analyses
\begin{align*}
S_x &= \frac{1}{n_x-1} \sum_{i=1}^{n_x} \tilde{x}_{i \cdot} \tilde{x}_{i \cdot}^\top = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top 
\end{align*}
thereby obtaining loadings $\eta_{j}^x$ and eigenvalues $\lambda^x_j$ for $j=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. The standardization implies that the diagonal elements of $S_x$, $S_y$, and $S_z$ all equal 1, and thus also that
\begin{equation*}
\sum_{j=1}^d \lambda_j^x = \sum_{j=1}^d \lambda_{j}^y = \sum_{j=1}^d \lambda_{j}^z =  d.
\end{equation*}
This identity will simplify some expressions below. We note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$. This implies that the loadings and the eigenvalues can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.

@@

\subsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the eigenvalues of the correlation matrices. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of eigenvalues, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1^z,\lambda_{1}^x-\lambda_{1}^y), \;
(\lambda_1^z + \lambda_2^z,\lambda_{1}^x+\lambda_{2}^x-\lambda_{1}^y-\lambda_{2}^y), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j^z, \sum_{j=1}^d \lambda_{j}^x - \sum_{j=1}^d \lambda_{j}^y \bigg)
\end{align*}
Due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the  first axis, the less alike the cumulative eigenvalue sums for the two datasets are. Moreover, a positive cumulative difference implies that dataset $\mathbf{X}$ holds more information in the first components than dataset $\mathbf{Y}$ does. By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more influential components drive the visual impression.

@@

In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k^z + \lambda_{k+1}^z}{2} {\bigg( \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg)}^2.
\end{align*}
We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_x$ and $n_y$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics. This should be done a large (e.g. 10000) number of times. Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
test statistics at least as large as %than
the one we found for the original datasets.

@@

\subsection*{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation*}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align*}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets can always be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

@@

\subsection*{The chroma plot}
The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align*}
\omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
\omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
\end{align*}
where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively. Due to the indeterminacy of the loading signs, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
\begin{align*}
\tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
\tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot). In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$. 
If there is eigenvalue indeterminancy for neighboring components, as discussed above, the loading patterns for these components will not have a meaningful interpretation. Therefore, one should not compare the chroma bars for component that were identified to suffer from eigenvalue indeterminancy by the angle plot.

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing}, and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in Appendix S3. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

\begin{table}[ht]
\centering
\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and 75 percentiles are listed in parentheses. The scales are constructed such that they all run from 0-10.} 
\label{tableDistr}

\begin{tabular}{lccc}
\hline
  & Denmark & Bulgaria & Sweden \\
\hline
Evaluative wellbeing    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
Emotional wellbeing     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
Community wellbeing     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
\hline
\end{tabular} 
\end{table}

@@

The chroma plot in Figure \ref{plotBG.pancake} allows us to look closer into these components. Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish. In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.

@@

Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components. We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets. In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and for instance conclude using Table~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally happier than the least happy people in Sweden. A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.

@@

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure2_v3.pdf}
\caption{CE plots for the simulated datasets. Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel. Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
\label{plot.simCE} 
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure3_v3.pdf}
\caption{Angle plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.}
\label{plot.simAngle}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure4_v3.pdf}
\caption{Chroma plots for the simulated datasets. Dataset $\mathbf{A}$ in top panel and dataset $\mathbf{B}$ in bottom panel. Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.}
\label{plot.simChroma}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure6_v3.pdf}
\caption{CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being. The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures. In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
\label{plotBG.cehair}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure7_v3.pdf}
\caption{Chroma plot for comparing Danish and Bulgarian data on psychological well-being. A chroma plot comparing the 2nd, 3rd, and 4th principal components of the Bulgarian- and Danish psychological well-being data. The component-bars are annotated with their relative variance contributions. }
\label{plotBG.pancake}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure8_v3.pdf}
\caption{A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being. The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
\label{plotSE.cehair}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure9_v3.pdf}
\caption{Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples. For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.}
\label{plotSE.pancake}
\end{figure}

>>>>

\usepackage{caption}
\captionsetup{font={stretch=2}}

@@

The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
\begin{equation}
\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}
\end{equation}
 be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
\begin{enumerate}
\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{\mathbf{X}}_{j} \in \RR^{n_x \times d}$ be the standardized dataset.
\item Form the principal component analyses
\begin{align}
S_x &= \frac{1}{n_x-1} \sum_{i=1}^{n_x} \tilde{x}_{i \cdot} \tilde{x}_{i \cdot}^\top = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top 
\end{align}
thereby obtaining loadings $\eta_{j}^x$ and eigenvalues $\lambda^x_j$ for $j=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. The standardization implies that the diagonal elements of $S_x$, $S_y$, and $S_z$ all equal 1, and thus also that
\begin{equation}
\sum_{j=1}^d \lambda_j^x = \sum_{j=1}^d \lambda_{j}^y = \sum_{j=1}^d \lambda_{j}^z =  d.
\end{equation}
This identity will simplify some expressions below. We note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$. This implies that the loadings and the eigenvalues can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.

@@

\subsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the eigenvalues of the correlation matrices. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of eigenvalues, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align}
&(0,0), \;
(\lambda_1^z,\lambda_{1}^x-\lambda_{1}^y), \;
(\lambda_1^z + \lambda_2^z,\lambda_{1}^x+\lambda_{2}^x-\lambda_{1}^y-\lambda_{2}^y), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j^z, \sum_{j=1}^d \lambda_{j}^x - \sum_{j=1}^d \lambda_{j}^y \bigg)
\end{align}
Due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the  first axis, the less alike the cumulative eigenvalue sums for the two datasets are. Moreover, a positive cumulative difference implies that dataset $\mathbf{X}$ holds more information in the first components than dataset $\mathbf{Y}$ does. By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more influential components drive the visual impression.

@@

In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
\begin{align}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k^z + \lambda_{k+1}^z}{2} {\bigg( \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg)}^2.
\end{align}
We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_x$ and $n_y$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics. This should be done a large (e.g. 10000) number of times. Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
test statistics at least as large as %than
the one we found for the original datasets.

@@

\subsection*{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets can always be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

@@

\subsection*{The chroma plot}
The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align}
\omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
\omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
\end{align}
where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively. Due to the indeterminacy of the loading signs, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
\begin{align}
\tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
\tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
\end{align}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot). In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$. 
If there is eigenvalue indeterminancy for neighboring components, as discussed above, the loading patterns for these components will not have a meaningful interpretation. Therefore, one should not compare the chroma bars for component that were identified to suffer from eigenvalue indeterminancy by the angle plot.

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative well-being}, \textit{Emotional well-being}, \textit{Functioning}, \textit{Vitality}, \textit{Community well-being}, and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in Appendix S3. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Figure~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

%\begin{figure}[ht]
%\centering
%\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and %75 percentiles are listed in parentheses. The scales are constructed such that they all run from 0-10.} 
%\label{tableDistr}
%
%\begin{tabular}{lccc}
%\hline
%  & Denmark & Bulgaria & Sweden \\
%\hline
%Evaluative well-being    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
%Emotional well-being     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
%Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
%Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
%Community well-being     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
%Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
%\hline
%\end{tabular} 
%\end{figure}

@@

The chroma plot in Figure \ref{plotBG.pancake} allows us to look closer into these components. Here, we find that the relative importance of the \textit{Community well-being} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish. In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.

@@

Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components. We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets. In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and for instance conclude using Figure~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally happier than the least happy people in Sweden. A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.

@@

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure2_AnneHelbyPetersen_v3.pdf}
\caption{CE plots for the simulated datasets. Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel. Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
\label{plot.simCE} 
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure3_AnneHelbyPetersen_v3.pdf}
\caption{Angle plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.}
\label{plot.simAngle}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure4_AnneHelbyPetersen_v3.pdf}
\caption{Chroma plots for the simulated datasets. Dataset $\mathbf{A}$ in top panel and dataset $\mathbf{B}$ in bottom panel. Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.}
\label{plot.simChroma}
\end{figure}

@@

\begin{figure}[H]
\centering
\caption{Median values of each of the six dimensions of psychological well-being, stratified by country. 25 and 75 percentiles are listed in parentheses. The scales are constructed such that they all run from 0-10.} 
\label{tableDistr}

\begin{tabular}{lccc}
\hline
  & Denmark & Bulgaria & Sweden \\
\hline
Evaluative well-being    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
Emotional well-being     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
Community well-being     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
\hline
\end{tabular} 
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure7_AnneHelbyPetersen_v3.pdf}
\caption{CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being. The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures. In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
\label{plotBG.cehair}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure8_AnneHelbyPetersen_v3.pdf}
\caption{Chroma plot for comparing Danish and Bulgarian data on psychological well-being. A chroma plot comparing the 2nd, 3rd, and 4th principal components of the Bulgarian- and Danish psychological well-being data. The component-bars are annotated with their relative variance contributions. }
\label{plotBG.pancake}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure9_AnneHelbyPetersen_v3.pdf}
\caption{A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being. The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
\label{plotSE.cehair}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.8]{Figure10_AnneHelbyPetersen_v3.pdf}
\caption{Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples. For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.}
\label{plotSE.pancake}
\end{figure}
